Working memory performance in adverse environments: Enhanced, impaired, or intact?

### 2024-06-25 manuscript/manuscript.qmd
Add submission of stage 2 manuscript to README

```{r include = FALSE}
library(flextable)
library(stringr)
library(dplyr)
library(tidyr)
library(tibble)
library(lavaan)
library(officer)

@@

```

@@

All scripts and materials needed to reproduce the findings are available on the article's Github repository ([https://stefanvermeent.github.io/liss_wm_profiles_2023/](https://stefanvermeent.github.io/liss_wm_profiles_2023/)).
We also include instructions on how to reproduce each step of our analyses, as well as simulated data to ensure computational reproducibility. 
In this paper, we make use of data from the LISS panel (Longitudinal Internet studies for the Social Sciences) managed by the non-profit research institute Centerdata (Tilburg University, the Netherlands). 
All datasets are available in the LISS data archive. 
Researchers who want to access the data are required to sign a statement confirming that information about individual persons, households, etc., will not be released to others (go to [https://statements.centerdata.nl](https://statements.centerdata.nl) for more information).

@@

# Proposal abstract

Decades of research have shown that adversity tends to be associated with lower working memory (WM) performance.
This literature has mainly focused on impairments in the capacity to hold information available in WM for further processing.
However, some recent adaptation-based studies suggest that certain types of adversity can leave intact, or even enhance, the ability to rapidly update information in WM.
One key challenge is that WM capacity and updating tasks tend to covary.
This is likely due to the fact that both types of tasks require the creation and maintenance of bindings in WM; links between mental representations of information in WM.
To estimate the associations between adversity and different processes in WM, we need to isolate variance in performance related to WM capacity from variance in performance related to updating ability.
In this Registered Report, we combine archived and newly collected data in the Dutch Longitudinal Internet studies for the Social Sciences (LISS) panel, which includes a representative sample of the Dutch adult population.
Participants completed three WM tasks: two complex span tasks and a task measuring both binding and updating of information.
In addition, we will estimate participants' exposure to neighborhood threat, material deprivation, and unpredictability.
Using structural equation modeling, we will estimate associations between the three types of adversity and latent estimates of WM capacity and updating.
These findings will advance our theoretical understanding of how adversity is associated with WM, which will aid interventions aimed at alleviating performance difficulties and leveraging areas of strength.

@@

With deficit theories focusing on WM capacity and adaptation-based theories on WM updating, we may wonder how capacity and updating are related to each other.
Performance on tasks measuring WM capacity and updating tend to be substantially correlated [in the range of .20-.50\; @frischkorn_2022; @loffler_2022].
This overlap appears to stem from shared demands of both types of tasks, in particular the need to create and maintain arbitrary bindings [@gruszka_2017; @oberauer_2009; @wilhelm_2013].
The term *binding* refers to the process of mapping memory items to specific positions in WM (e.g., serial, spatial, or temporal positions, depending on the task) [@oberauer_2009; @oberauer_2019].
For example, on most WM tasks, correct recall of memory items depends on remembering them in their correct serial position, or in relation to the location where they were presented.

@@

In this study, we will estimate associations between latent estimates of WM capacity and updating with three types of adversity: threat, deprivation, and unpredictability. 
Together, these adversity types capture key dimensions in contemporary models of adversity [@ellis_2009; @ellis_2022; @mclaughlin_2016; @mclaughlin_2021]. 
Threat refers to experiences involving the potential for harm imposed by others. 
We focus on perceived neighborhood violence, the extent to which an individual reports having been exposed to acts of violence in their neighborhood. 
Deprivation refers to having a low level of resources. 
We focus on perceived material deprivation, a (perceived) lack of access to material resources. 
Unpredictability refers to variation in material deprivation over time. 
This definition is inspired by, but deviates from the harshness-unpredictability framework, in which unpredictability is defined as stochastic variation in harshness (age-specific rates in morbidity and mortality) over space and time [@ellis_2009; @ellis_2022].
We will not calculate unpredictability in neighborhood threat given that participants have at most six timepoints, and often as few as one or two, which is insufficient to calculate variation over time [@walasek_2024].

We will address three research questions.
First, what is the association between adversity and WM capacity?
Second, what is the association between adversity and WM updating *after* accounting for WM capacity?
Third, are the directions and strengths of these associations similar or different for neighborhood threat, material deprivation, and unpredictability?

We will evaluate evidence for deficit- and adaptation-based frameworks (see Figure 1A for a visual summary, and Appendix 1 for the study design plan). 
Crucially, as deficit and adaptation processes can operate in concert [@frankenhuis_2020], we could find support (or lack thereof) for both frameworks in the same model. 
We distinguish between three between-person data patterns: (1) lowered performance, (2) enhanced performance, and (3) intact performance.
We define lowered performance as a statistically significant negative association between a type of adversity and WM capacity or updating (irrespective of effect size).
We define enhanced performance as a statistically significant positive association between a type of adversity and WM capacity or updating (irrespective of effect size).
We define intact performance as an association between a type of adversity and WM capacity or updating that has a standardized effect smaller than 0.1 *and* larger than -0.1---even if the effect is statistically different from zero---which we will test using Two One-Sided T-Tests (TOST) equivalence testing [see the 'Primary analyses' section\; @lakens_2018].

@@

We also have two expectations based on prior studies. 
First, we expect the association between material deprivation and WM capacity to be more negative than the associations with unpredictability and neighborhood threat. 
This follows from findings showing that cognitive abilities are more negatively associated with cognitive deprivation than threat [@salhi_2021; @sheridan_2020].
Although cognitive and material deprivation are distinct types of deprivation, they tend to be correlated, and are both associated with limited access to resources that stimulate cognitive development and functioning [@bradley_2001; @lurie_2024; @rosen_2019].
Therefore, we expect that their associations with WM will have comparable effect sizes. 
Second, researchers have hypothesized that WM updating is particularly adaptive in unpredictable and threatening environments, as it facilitates keeping track of unpredictable changes and sudden threats. 
Therefore, we expect WM updating to be associated with unpredictability and neighborhood threat, but not with material deprivation [@young_2018; but see @young_2022].

@@

```{r}
#| label: Figure1
#| fig-width: 6.5
#| fig-height: 7
#| dpi: 600
#| out-width: 6in
#| fig-cap: | 
#|   **Figure 1.** Overview of predictions derived from deficit and adaptation frameworks. Panel A depicts the most likely between-person data patterns based on previous literature, and whether we would consider them consistent with deficit and adaptation frameworks (see the main text for more details). Panel B depicts an overview of the Structural Equation Model. Ellipses represent latent variables, rectangles represent manifest variables, and circles represent residual variances. Unidirectional solid lines represent factor loadings, bidirectional solid lines represent covariances, and dashed lines represent regression paths. All four manifest WM measures load on a latent WM capacity factor, reflecting the fact that people have to hold information active in WM on all tasks. We fix the loading of WM capacity on the Binding Task to 1, reflecting the idea that the ability to create and maintain bindings is the main limiting factor in WM capacity [@gruszka_2017; @oberauer_2009; @wilhelm_2013]. WM updating is modeled as a latent factor capturing the residual variance in the updating task after accounting for variance related to WM capacity. INR = income-to-needs ratio; Perc. Scarcity = perceived scarcity; SD = standard deviation.
knitr::include_graphics("fig/fig1.png")
```

@@

Our study will include 800 participant who were randomly sampled from the Longitudinal Internet studies for the Social Sciences (LISS) panel [@scherpenzeel_2011].
The LISS panel is a representative probability sample of roughly 5,000 Dutch households (~7,500 individuals) drawn from the population register by Statistics Netherlands on an invite-only basis.
Households without a computer or internet connection are provided with these facilities by LISS.
Each year, participants complete the same core battery of questionnaires about---among other topics---their financial situation in the past year.
In addition, participants can complete additional online questionnaires every month, with variable content.
The current study integrates two data sources.
First, our sample of 800 participants participated in a new LISS study between October 2023 and February 2024 (hereafter referred to as 'newly collected data'), in which we included a measure of neighborhood threat and multiple measures of working memory.
Second, we will access data that were previously collected in LISS (hereafter referred to as 'the LISS archive').
See Figure 2 for a visual overview of the data sources and their measurement timepoints.
We signed a contract with LISS stipulating that we will receive access to the newly collected data only after Stage 1 acceptance of this Registered Report.

```{r}
#| label: Figure2
#| fig-width: 6.5
#| fig-height: 7
#| dpi: 600
#| out-width: 6in
#| fig-cap: | 
#|   **Figure 2.**. Overview of the different data sources used in this study. We distinguish between measures taken from the LISS data archive and measures that were newly collected in our own study between October 2023 and February 2024. Perceived scarcity and income were collected yearly in the full panel from 2008 – 2023. Neighborhood crime and crime victimization were collected across six waves between 2008 and 2018. In the newly collected data, we collected data on a measure of neighborhood threat and multiple measures of working memory. Note that participants may not have data across all timepoints of the archived studies because they joined the LISS panel more recently or because they did not participate in each wave.
knitr::include_graphics("fig/fig2.png")
```

We based our power analysis on simulations reported by @kretzschmar_2019, determining the required sample size to detect a small effect size ($\beta$ = 0.1) with at least 90% power at $\alpha$ = 0.05.
Assuming a reliability of at least 0.7 [which is typical for WM tasks with a number of trials similar to ours\; e.g., @wilhelm_2013], we would require a sample size of *N* = 730.
Anticipating some exclusions, we decided to include 800 participants.
Participants were eligible for inclusion if they 1) were currently between 18 and 55 years old, 2) had completed at least one wave of an archived longitudinal LISS study containing measures that we use to operationalize crime neighborhood threat (see below), and 3) had given permission to link their LISS data to government microdata (not relevant here).

To ensure sufficient representation of people from lower socioeconomic backgrounds, *half* the total sample was sampled from participants who reported one or more of the following at least once in the three years: (1) a monthly income < €1,500, (2) HAVO or VWO as highest completed education (which are the two highest levels in Dutch secondary education), or (3) a score of 4 or lower on the 'ladder of life' ("If you imagine a ‘ladder of life’, where the first step represents the worst possible life, and the tenth (top) step the best possible life, on what step would you place yourself?").
Participants will be excluded if they (1) switched to and interacted with other browser tabs *during* one or more of the cognitive tasks, (2) did not perform above chance level.

@@

@@

**Perceived neighborhood crime.** We included four items from the LISS archive collected across six waves ([https://doi.org/10.17026/dans-zch-j8xt](https://doi.org/10.17026/dans-zch-j8xt)), in which participants answered how often it happens that they 1) "avoid certain areas in your place of residence because you perceive them as
unsafe", 2) "do not respond to a call at the door because you feel that it is unsafe", 3) "leave valuable items at home to avoid theft or robbery in the street?", 4) "make a detour, by car or on foot, to avoid unsafe areas?" on a scale of 1 ("(Almost) never"), 2 ("Sometimes"), or 3 ("Often").
We will recode these items so that 0 indicates "(Almost) never". 
We then sum the responses within each wave for which participants have data, and calculate an average across the waves.

In addition, we implemented the Neighborhood Violence Scale [NVS\; @frankenhuis_2018; @frankenhuis_2020] in the newly collected data.
The NVS includes seven items measuring perceived exposure to neighborhood violence (e.g., "Crime is common in the neighborhood where I live"; "Where I live, it is important to be able to defend yourself against physical harm").
Participants answered these questions on a scale of 1 ("Completely disagree") to 7 ("Completely agree").
We will compute an average of the seven items.

**Crime victimization.** We used data from the LISS archive collected across six waves (same dataset as above), in which participants indicated whether they fell victim to eight types of crime over the two years prior to a particular wave (0 = no, 1 = yes).
We included seven items concerning exposure to crime: (1) burglary or attempted burglary; (2) theft from their car; (3) theft of their wallet or purse, handbag, or other personal possession; (4) wreckage of their car or other private property; (5) intimidation by any other means; (6) maltreatment of such serious nature that it required medical attention; (7) maltreatment that did not require medical attention.
We compute a variety score by summing the exposures to *unique* types of crime across all waves.
Thus, if a participants reports exposure to the same type of crime on separate waves, this will count as one exposure in the total score [@sweeten_2012].

**Neighborhood threat composite.** We will first compute an average across time for each measure separately (i.e., the two measures of neighborhood crime and the measure of crime victimization).
If all correlations between averaged measures are equal to or larger than .60 (i.e., indicating a "strong" correlation), then we will compute a uniformly weighted average.
If the correlation is lower than .60, we will apply Principal Component Analysis (PCA) to the averaged measures and extract only the first principal component score.

@@

We will measure material deprivation with two separate indicators: perceived scarcity and the income-to-needs ratio.

**Perceived scarcity (mean).** We will use a few items from the LISS archive that were collected on a yearly basis between 2008 and 2023 ([https://doi.org/10.57990/1gr4-bf42](https://doi.org/10.57990/1gr4-bf42)) to index perceived scarcity.
First, participants indicated how hard or easy it currently is to live off the income of their household, on a scale from 0 (very hard) to 10 (very easy).
Second, participants were asked to choose which of the following best applied to their current situation: (1) "we are accumulating debt"; (2) "we are somewhat eating into savings"; (3) "we are just managing to make ends meet"; (4) "we have a little bit of money to spare"; (5) "we have a lot of money to spare".
Responses will be reverse-coded, so that higher scores indicate a worse financial situation.
Third, participants answered which of the following issues they were confronted with at present (0 = no, 1 = yes): (1) "having trouble making ends meet"; (2) unable to quickly replace things that break"; (3) "having to lend money for necessary expenditures"; (4) "running behind in paying rent/mortgage or general utilities"; (5) "debt collector/bailiff at the door in the last month"; (6) "received financial support from family or friends in the last month".

We will first compute the average across time for each item separately, and examine correlation between the item averages.
We will follow the same guidelines as with neighborhood threat to determine if we will compute a uniformly weighted average or extract the first principal component using PCA.

**Income-to-needs (mean).** We will calculate an income-to-needs ratio for each year using monthly self-reported net household income from the LISS archive ([https://doi.org/10.57990/qn3k-as78](https://doi.org/10.57990/qn3k-as78)).
Zero values in household income will be set to missing, as these could either indicate the lack of an income or an unwillingness to disclose the income.
If monthly household income is missing (or zero) for an entire year for a participant, we will use, if available, the yearly net household income they reported in the LISS archive ([https://doi.org/10.57990/1gr4-bf42](https://doi.org/10.57990/1gr4-bf42)), dividing it by 12 to obtain a monthly estimate.
First, we will divide the average income per year by the *poverty threshold*, as determined by Statistics Netherlands (CBS) [@brakel_2023\; CBS, personal communication, December 15, 2023].
As thresholds are only provided for households with up to three children, we will apply the threshold of a household with three children to households with more than three children.
Likewise, we will apply the threshold of a household with two adults for households that contain three or more adults.
Second, we will calculate the average within-person income-to-needs ratio for each year by averaging across the monthly income-to-needs estimates.
We will reverse code the yearly income-to-needs ratio so that higher scores indicate more deprivation.

@@

**Perceived scarcity (SD/mean).** This measure is based on the same items as outlined above (see Perceived scarcity (mean)).
We will compute unpredictability over time in perceived scarcity using the coefficient of variation, which is the within-person standard deviation across years divided by the mean [@key_2017; @liu_2022; @ugarte_2023; @walasek_2024; @young_2020].
The mean and standard deviation in income have been found to be strongly negatively correlated, indicating that people with lower incomes tend to experience less variability in income [@li_2018; @young_2023].
For that reason, the standard deviation alone has been called into question as a measure of adversity, as the same fluctuation in income can have a greater relative impact for people close to the poverty line than for people with high incomes.

To compute an overall measure of unpredictability in perceived scarcity, we will follow the same procedure as with mean perceived scarcity, but based on the coefficient of variation.

**Income-to-needs (SD/mean).** Similar to perceived scarcity, we will compute unpredictability over time in the income-to-needs ratio using the coefficient of variation.

@@

Participants first practice the letter task (three times), then the math task (eight times), and then the full task (three times).
If they perform at or below chance, they have the opportunity to either repeat a part or advance to the next part.
After practicing, participants complete 9 test trials, with a total of 45 recall items and 45 math items.
We will compute an operation span score by calculating the proportion of letters recalled in the correct sequential position across trials [@conway_2005].

**Rotation Span Task.** The Rotation Span Task (Figure 2B) is similar to the Operation Span Task and is adopted from @wilhelm_2013.
On each trial, the task is to memorize the orientation of a sequence of arrows in the correct order.
Arrows can take on eight different orientations, with increments of 45$\circ$.
Each arrow is presented for 1,000 ms in the center of the screen. 
Next, participants see a capital 'G' or 'F' that is rotated at one of eight different orientations, with increments of 45$\circ$.
Their task is to indicate whether the letter is mirrored or not.
On each trial, participants have to memorize between two to five arrows, with each set size repeated three times.
At the end of each sequence, all arrows are presented simultaneously, and participants click the arrows in the correct order.

>>>>

```{r include = FALSE}
library(flextable)
library(stringr)
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(lavaan)
library(officer)

@@

clean_data <- read_csv('../data/liss_data/clean_data.csv')
load("../analysis_objects/confirmatory_results.RData") # SEM results
load("../analysis_objects/exploratory_results.RData") # SEM results
load("../analysis_objects/pca.RData") # PCA results for threat and unpredictability in perceived scarcity
load("../analysis_objects/tables.RData")
load("../analysis_objects/fig5.RData")
load("../analysis_objects/fig6.RData")
```

@@

All scripts and materials needed to reproduce the findings are available on the article's Github repository ([https://stefanvermeent.github.io/liss_wm_profiles_2023/](https://stefanvermeent.github.io/liss_wm_profiles_2023/)).
We also include instructions on how to reproduce each step of our analyses.
In this paper, we make use of data from the LISS panel (Longitudinal Internet studies for the Social Sciences) managed by the non-profit research institute Centerdata (Tilburg University, the Netherlands). 
All datasets are available in the LISS data archive. 
Researchers who want to access the data are required to sign a statement confirming that information about individual persons, households, etc., will not be released to others (go to [https://statements.centerdata.nl](https://statements.centerdata.nl) for more information).

@@

# Abstract

Decades of research have shown that adversity tends to be associated with lower working memory (WM) performance.
This literature has mainly focused on impairments in the capacity to hold information available in WM for further processing.
However, some recent adaptation-based studies suggest that certain types of adversity can leave intact, or even enhance, the ability to rapidly update information in WM.
One key challenge is that WM capacity and updating tasks tend to covary, as both types of tasks require the creation and maintenance of bindings in WM; links between mental representations of information in WM.
To estimate the associations between adversity and different processes in WM, we need to isolate variance in performance related to WM capacity from variance in performance related to updating ability.
In this Registered Report, participants from the Dutch Longitudinal Internet studies for the Social Sciences (LISS) panel completed three WM tasks: two complex span tasks and a task measuring both binding and updating of information.
In addition, we estimated participants' exposure to neighborhood threat, material deprivation, and unpredictability.
We estimated associations between the three types of adversity and latent estimates of WM capacity and updating using structural equation modeling.
We did not find consistent associations between adversity and WM capacity or updating, nor did we find evidence that the associations were practically equivalent to zero.
Our results show that adversity researchers should account for overlap in WM tasks when estimating specific WM abilities.

@@

With deficit theories focusing on WM capacity and adaptation-based theories on WM updating, we may wonder how capacity and updating are related to each other.
Performance on tasks measuring WM capacity and updating tend to be substantially correlated [in the range of .20-.50\; @frischkorn_2022; @loffler_2024].
This overlap appears to stem from shared demands of both types of tasks, in particular the need to create and maintain arbitrary bindings [@gruszka_2017; @oberauer_2009; @wilhelm_2013].
The term *binding* refers to the process of mapping memory items to specific positions in WM (e.g., serial, spatial, or temporal positions, depending on the task) [@oberauer_2009; @oberauer_2019].
For example, on most WM tasks, correct recall of memory items depends on remembering them in their correct serial position, or in relation to the location where they were presented.

@@

In this study, we estimated associations between latent estimates of WM capacity and updating with three types of adversity: threat, deprivation, and unpredictability. 
Together, these adversity types capture key dimensions in contemporary models of adversity [@ellis_2009; @ellis_2022; @mclaughlin_2016; @mclaughlin_2021]. 
Threat refers to experiences involving the potential for harm imposed by others. 
We focused on perceived neighborhood violence, the extent to which an individual reports having been exposed to acts of violence in their neighborhood. 
Deprivation refers to having a low level of resources. 
We focused on perceived material deprivation, a (perceived) lack of access to material resources. 
Unpredictability refers to variation in material deprivation over time. 
This definition is inspired by, but deviates from the harshness-unpredictability framework, in which unpredictability is defined as stochastic variation in harshness (age-specific rates in morbidity and mortality) over space and time [@ellis_2009; @ellis_2022].
We did not calculate unpredictability in neighborhood threat given that participants had at most six timepoints, and often as few as one or two, which is insufficient to calculate variation over time [@walasek_2024].

We addressed three research questions.
First, what is the association between adversity and WM capacity?
Second, what is the association between adversity and WM updating *after* accounting for WM capacity?
Third, are the directions and strengths of these associations similar or different for neighborhood threat, material deprivation, and unpredictability?

We evaluated evidence for deficit- and adaptation-based frameworks (see Figure 1A for a visual summary, and Appendix 1 for the study design plan). 
Crucially, as deficit and adaptation processes can operate in concert [@frankenhuis_2020], we could find support (or lack thereof) for both frameworks in the same model. 
We distinguished between three between-person data patterns: (1) lowered performance, (2) enhanced performance, and (3) intact performance.
We defined lowered performance as a statistically significant negative association between a type of adversity and WM capacity or updating (irrespective of effect size).
We defined enhanced performance as a statistically significant positive association between a type of adversity and WM capacity or updating (irrespective of effect size).
We defined intact performance as an association between a type of adversity and WM capacity or updating that has a standardized effect smaller than 0.1 *and* larger than -0.1---even if the effect is statistically different from zero---which we tested using Two One-Sided T-Tests (TOST) equivalence testing [see the 'Primary analyses' section\; @lakens_2018].

@@

We also had two expectations based on prior studies. 
First, we expected the association between material deprivation and WM capacity to be more negative than the associations with unpredictability and neighborhood threat. 
This follows from findings showing that cognitive abilities are more negatively associated with cognitive deprivation than threat [@salhi_2021; @sheridan_2020].
Although cognitive and material deprivation are distinct types of deprivation, they tend to be correlated, and are both associated with limited access to resources that stimulate cognitive development and functioning [@bradley_2001; @lurie_2024; @rosen_2019].
Therefore, we expected that their associations with WM would have comparable effect sizes. 
Second, researchers have hypothesized that WM updating is particularly adaptive in unpredictable and threatening environments, as it facilitates keeping track of unpredictable changes and sudden threats. 
Therefore, we expected WM updating to be associated with unpredictability and neighborhood threat, but not with material deprivation [@young_2018; but see @young_2022].

@@

```{r}
#| label: Figure1
#| fig-width: 6.5
#| fig-height: 7
#| dpi: 600
#| out-width: 6in
#| fig-cap: | 
#|   **Figure 1.** Overview of predictions derived from deficit and adaptation frameworks. Panel A depicts the most likely between-person data patterns based on previous literature, and whether we would consider them consistent with deficit and adaptation frameworks (see the main text for more details). Panel B depicts an overview of the preregistered Structural Equation Model. Note that this model differs slightly from the final model (see Figure 4). Ellipses represent latent variables, rectangles represent manifest variables, and circles represent residual variances. Unidirectional solid lines represent factor loadings, bidirectional solid lines represent covariances, and dashed lines represent regression paths. All four manifest WM measures loaded on a latent WM capacity factor, reflecting the fact that people have to hold information active in WM on all tasks. We fixed the loading of WM capacity on the Binding Task to 1, reflecting the idea that the ability to create and maintain bindings is the main limiting factor in WM capacity [@gruszka_2017; @oberauer_2009; @wilhelm_2013]. WM updating was modeled as a latent factor capturing the residual variance in the updating task after accounting for variance related to WM capacity. INR = income-to-needs ratio; Perc. Scarcity = perceived scarcity; SD = standard deviation.
knitr::include_graphics("fig/fig1.png")
```

@@

Our study included 800 participant who were randomly sampled from the Longitudinal Internet studies for the Social Sciences (LISS) panel [@scherpenzeel_2011].
The LISS panel is a representative probability sample of roughly 5,000 Dutch households (~7,500 individuals) drawn from the population register by Statistics Netherlands on an invite-only basis.
Households without a computer or internet connection are provided with these facilities by LISS.
Each year, participants complete the same core battery of questionnaires about---among other topics---their financial situation in the past year.
In addition, participants can complete additional online questionnaires every month, with variable content.
The current study integrated two data sources.
First, our sample of 800 participants participated in a new LISS study between October 2023 and February 2024 (hereafter referred to as 'newly collected data'), in which we included a measure of neighborhood threat and multiple measures of working memory.
Second, we accessed data that were previously collected in LISS (hereafter referred to as 'the LISS archive').
See Figure 2 for a visual overview of the data sources and their measurement timepoints.
We signed a contract with LISS stipulating that we would receive access to the newly collected data only after Stage 1 acceptance of this Registered Report.

```{r}
#| label: Figure2
#| fig-width: 6.5
#| fig-height: 7
#| dpi: 600
#| out-width: 6in
#| fig-cap: | 
#|   **Figure 2.**. Overview of the different data sources used in this study. We distinguished between measures taken from the LISS data archive and measures that were newly collected in our own study between October 2023 and February 2024. Perceived scarcity and income were collected yearly in the full panel from 2008 – 2023. Neighborhood crime and crime victimization were collected across six waves between 2008 and 2018. In the newly collected data, we collected data on a measure of neighborhood threat and multiple measures of working memory. Note that participants did not have data across all timepoints of the archived studies because they joined the LISS panel more recently or because they did not participate in each wave.
knitr::include_graphics("fig/fig2.png")
```

We based our power analysis on simulations reported by @kretzschmar_2019, determining the required sample size to detect a small effect size ($\beta$ = 0.1) with at least 90% power at $\alpha$ = 0.05.
Assuming a reliability of at least 0.7 [which is typical for WM tasks with a number of trials similar to ours\; e.g., @wilhelm_2013], we required a sample size of *N* = 730.
Anticipating some exclusions, we decided to include 800 participants.
Participants were eligible for inclusion if they 1) were currently between 18 and 55 years old, 2) had completed at least one wave of an archived longitudinal LISS study containing measures that we use to operationalize crime neighborhood threat (see below), and 3) had given permission to link their LISS data to government microdata (not relevant here).

To ensure sufficient representation of people from lower socioeconomic backgrounds, *half* the total sample was sampled from participants who reported one or more of the following at least once in the three years: (1) a monthly income < €1,500, (2) HAVO or VWO as highest completed education (which are the two highest levels in Dutch secondary education), or (3) a score of 4 or lower on the 'ladder of life' ("If you imagine a ‘ladder of life’, where the first step represents the worst possible life, and the tenth (top) step the best possible life, on what step would you place yourself?").
Participants were excluded if they (1) switched to and interacted with other browser tabs *during* one or more of the cognitive tasks, (2) did not perform above chance level on the secondary processing tasks.
The final sample consisted of `r nrow(clean_data)` participants.

```{r}
table1 |> 
  flextable::autofit() |> 
  add_header_row(
    values = " ",
    colwidths = 2
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 1. "), "Descriptive statistics."),
    part = "header"
  ) |> 
  border_remove() |> 
  border(i = 2, border.top = fp_border_default(), part = "header") |>  
  border(i = 2, border.bottom = fp_border_default(), part = "header") |> 
  border(i = 15, border.bottom = fp_border_default(), part = "body") 
```

@@

All independent variables, except for the income-to-needs ratio (INR) consisted of multiple items and/or scales.
If all correlations between the items/scales were equal to or larger than .60 (i.e., indicating a "strong" correlation), then we computed a uniformly weighted average.
If the correlation was lower than .60, we applied Principal Component Analysis (PCA) to the averaged measures and extracted only the first principal component score.
We present bivariate correlations in Table 2, and histograms for all independent measures in the supplemental materials.

@@

**Perceived neighborhood crime.** We included four items from the LISS archive collected across six waves ([https://doi.org/10.17026/dans-zch-j8xt](https://doi.org/10.17026/dans-zch-j8xt)), in which participants answered how often it happens that they 1) "avoid certain areas in your place of residence because you perceive them as
unsafe", 2) "do not respond to a call at the door because you feel that it is unsafe", 3) "leave valuable items at home to avoid theft or robbery in the street?", 4) "make a detour, by car or on foot, to avoid unsafe areas?" on a scale of 1 ("(Almost) never"), 2 ("Sometimes"), or 3 ("Often").
We recoded these items so that 0 indicated "(Almost) never". 
We then summed the responses within each wave for which participants had data, and calculated an average across the waves.

In addition, we implemented the Neighborhood Violence Scale [NVS\; @frankenhuis_2018; @frankenhuis_2020] in the newly collected data.
The NVS includes seven items measuring perceived exposure to neighborhood violence (e.g., "Crime is common in the neighborhood where I live"; "Where I live, it is important to be able to defend yourself against physical harm").
Participants answered these questions on a scale of 1 ("Completely disagree") to 7 ("Completely agree").
We computed an average of the seven items.

**Crime victimization.** We used data from the LISS archive collected across six waves (same dataset as above), in which participants indicated whether they fell victim to eight types of crime over the two years prior to a particular wave (0 = no, 1 = yes).
We included seven items concerning exposure to crime: (1) burglary or attempted burglary; (2) theft from their car; (3) theft of their wallet or purse, handbag, or other personal possession; (4) wreckage of their car or other private property; (5) intimidation by any other means; (6) maltreatment of such serious nature that it required medical attention; (7) maltreatment that did not require medical attention.
We computed a variety score by summing the exposures to *unique* types of crime across all waves.
Thus, if a participants reported exposure to the same type of crime on separate waves, this counted as one exposure in the total score [@sweeten_2012].

**Neighborhood threat composite.** We first computed an average across time for each measure separately (i.e., the two measures of neighborhood crime and the measure of crime victimization).
Because correlations were below .60 (see Table 2), we then used PCA to extract only the first principal component score (R^2 = .20).
The threat component was most strongly determined by the NVS (`r pca_threat_fit$loadings[3] |> formatC(digits = 2, width = 2, format = 'f')`), followed by the perceived neighborhood crime scale from the LISS archive (`r pca_threat_fit$loadings[1] |> formatC(digits = 2, width = 2, format = 'f')`) and crime victimization (`r pca_threat_fit$loadings[2] |> formatC(digits = 2, width = 2, format = 'f')`).

@@

We measured material deprivation with two separate indicators: perceived scarcity and the income-to-needs ratio.

**Perceived scarcity (mean).** We used a few items from the LISS archive that were collected on a yearly basis between 2008 and 2023 ([https://doi.org/10.57990/1gr4-bf42](https://doi.org/10.57990/1gr4-bf42)) to index perceived scarcity.
First, participants indicated how hard or easy it currently is to live off the income of their household, on a scale from 0 (very hard) to 10 (very easy).
Second, participants were asked to choose which of the following best applied to their current situation: (1) "we are accumulating debt"; (2) "we are somewhat eating into savings"; (3) "we are just managing to make ends meet"; (4) "we have a little bit of money to spare"; (5) "we have a lot of money to spare".
Responses were reverse-coded, so that higher scores indicated a worse financial situation.
Third, participants answered which of the following issues they were confronted with at present (0 = no, 1 = yes): (1) "having trouble making ends meet"; (2) unable to quickly replace things that break"; (3) "having to lend money for necessary expenditures"; (4) "running behind in paying rent/mortgage or general utilities"; (5) "debt collector/bailiff at the door in the last month"; (6) "received financial support from family or friends in the last month".

We first computed the average across time for each item separately.
Because correlations were all above .60, we calculated a uniformly weighted average.

**Income-to-needs (mean).** We calculated an income-to-needs ratio for each year using monthly self-reported net household income from the LISS archive ([https://doi.org/10.57990/qn3k-as78](https://doi.org/10.57990/qn3k-as78)).
Zero values in household income were set to missing, as these could either indicate the lack of an income or an unwillingness to disclose the income.
If monthly household income is missing (or zero) for an entire year for a participant, we used, if available, the yearly net household income they reported in the LISS archive ([https://doi.org/10.57990/1gr4-bf42](https://doi.org/10.57990/1gr4-bf42)), dividing it by 12 to obtain a monthly estimate.
First, we divided the average income per year by the *poverty threshold*, as determined by Statistics Netherlands (CBS) [@brakel_2023\; CBS, personal communication, December 15, 2023].
As thresholds are only provided for households with up to three children, we applied the threshold of a household with three children to households with more than three children.
Likewise, we applied the threshold of a household with two adults for households that contained three or more adults.
Second, we calculated the average within-person income-to-needs ratio for each year by averaging across the monthly income-to-needs estimates.

@@

**Perceived scarcity (SD/mean).** This measure was based on the same items as outlined above (see Perceived scarcity (mean)).
We computed unpredictability over time in perceived scarcity using the coefficient of variation, which is the within-person standard deviation across years divided by the mean [@key_2017; @liu_2022; @ugarte_2023; @walasek_2024; @young_2020].
The mean and standard deviation in income have been found to be strongly negatively correlated, indicating that people with lower incomes tend to experience less variability in income [@li_2018; @young_2023].
For that reason, the standard deviation alone has been called into question as a measure of adversity, as the same fluctuation in income can have a greater relative impact for people close to the poverty line than for people with high incomes.

We first computed the standard deviation across time for each item separately.
because correlations were below .60 (see Table 2), we then used PCA to extract only the first principal component score (R^2 = .38).
The perceived unpredictability component was almost fully determined by the item about people's current situation (`r pca_p_scar_fit$loadings[3] |> formatC(digits = 2, width = 2, format = 'f')`), followed by difficulties to live off income (`r pca_p_scar_fit$loadings[2] |> formatC(digits = 2, width = 2, format = 'f')`) and financial troubles (`r  pca_p_scar_fit$loadings[1] |> formatC(digits = 2, width = 2, format = 'f')`).

**Income-to-needs (SD/mean).** Similar to perceived scarcity, we computed unpredictability over time in the income-to-needs ratio using the coefficient of variation.

```{r}
table2 |> 
  flextable() |> 
  flextable::set_header_labels(Variable = "") |> 
  flextable::autofit() |> 
  border_remove() %>% 
  border(i = 1, border.top = fp_border_default(), part = "header") %>% 
  border(i = 1, border.bottom = fp_border_default(), part = "header") %>% 
  border(i = 20, border.bottom = fp_border_default(), part = "body") %>% 
  add_header_row(
    values = " ",
    colwidths = 15
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 2. "), "Spearman correlation between the main independent variables."),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 15
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_i("Note: "), "* = ", as_i("p"), " < .05, ** = ", as_i("p"), " < .01, *** = ", as_i("p"), " < .001. ", "CV = coefficient of variance, INR = income-to-needs ratio, M = mean, Perc. Scarcity = perceived scarcity"), 
    part = "footer"
  ) |> 
  autofit()
```

@@

Participants first practiced the letter task (three times), then the math task (eight times), and then the full task (three times).
If they performed at or below chance, they had the opportunity to either repeat a part or advance to the next part.
After practicing, participants completed 9 test trials, with a total of 45 recall items and 45 math items.
We computed an operation span score by calculating the proportion of letters recalled in the correct sequential position across trials [@conway_2005].

**Rotation Span Task.** The Rotation Span Task (Figure 2B) is similar to the Operation Span Task and was adopted from @wilhelm_2013.
On each trial, the task is to memorize the orientation of a sequence of arrows in the correct order.
Arrows can take on eight different orientations, with increments of 45$\circ$.
Each arrow is presented for 1,000 ms in the center of the screen. 
Next, participants see a capital 'G' or 'F' that is rotated at one of eight different orientations, with increments of 45$\circ$.
Their task is to indicate whether the letter is mirrored or not.
On each trial, participants have to memorize between two to five arrows, with each set size repeated three times.
At the end of each sequence, all arrows are presented simultaneously, and participants click the arrows in the correct order.
